An Efficient Minibatch Acceptance Test for Metropolis-Hastings

### 2017-03-02 uai2017/paper.tex
Whew, logistic regression done.

% Daniel: NOTE, I'm calling temperature "K" so that "T" stays reserved for
% "number of samples of \theta we draw".
In Sections~\ref{ssec:gaussians} and~\ref{ssec:logistic}, we benchmark {\sc
MHminibatch} against the test from~\citep{icml2014c1_bardenet14} ({\sc
MHSubLhd}) and from~\citep{cutting_mh_2014} ({\sc AustereMH(c)}, and {\sc
AustereMH(nc)}). The latter two tests use a grid-search over minibatch sizes $m$
and per-test thresholds $\epsilon$ described in
Appendix~\ref{sssec:grid_search}. Throughout our descriptions, we refer to a
\emph{trial} as the period when an algorithm collects all its desired samples
$\{\theta_1,\ldots,\theta_T\}$, generally with $T=3000$ or $T=5000$.

@@

Figure~\ref{fig:gauss_mix_2} shows that {\sc MHminibatch} dominates in terms of
speed and efficiency. The histograms of the (final) minibatch sizes used each
iteration show that our method consumes significantly less data; the
distribution is short-tailed and the mean is 210, more than an order of
magnitude better compared to the other two methods (averages are 15562 and
16857). \textbf{Daniel: Xinlei, you should change these numbers as you see fit,
and also describe the number of trials you ran.} The sizes correspond to the
running times of the methods, excluding the likelihood computation of all data
points for {\sc MHSubLhd}, which would drastically increase running time.

@@

% Daniel: moving this here so figures don't overflow to next pages. We can tune
% the positioning if needed.
\begin{figure*}[t]
	\centering
	\includegraphics[width=1.0\linewidth]{logistic_regression_error_regions.png}
	\caption{
    Binary classification accuracy of the MCMC methods on the 1s vs 7s logistic
    regression task for MNIST-13k (left plot) and MNIST-100k (middle plot) as a
    function of cumulative data usage.  The right plot reports performance of
    {\sc MHminibatch} on both datasets when the temperature starts at $K=100$
    and drops to $K=1$ after a ``burn-in'' period of 200 samples of $\theta$.
    For all three plots, one standard deviation is indicated by the shaded error
    regions.
    }
	\label{fig:logistic_performance}
\end{figure*}

\begin{figure*}[t]
	\centering
    \includegraphics[width=0.9\linewidth]{logistic_regression_histograms_mnist8m.png}
	\caption{
    Minibatch sizes for a representative trial of logistic regression on
    MNIST-100k (similar to Figure~\ref{fig:logistic_histograms_mnist}). Both
    axes are on a log scale and have the same ranges across the three
    histograms.  See Section~\ref{ssec:logistic} for details.
    }
	\label{fig:logistic_histograms_mnist8m}
\end{figure*}

\begin{table}[t]
\caption{
Average minibatch sizes ($\pm$ one standard deviation) for the methods on the
logistic regression tasks of binary classification on MNIST-13k and MNIST-100k.
The averages are taken over 10 independent trials (5000 samples each) for
MNIST-13k and 5 independent trials (3000 samples each) for MNIST-100k.
}
\small % Daniel: yeah, sorry, but otherwise it doesn't fit in a column
\label{tab:logistic}
\vskip 0.15in
\begin{center}
\begin{tabular}{l l l l}
\textbf{Method/Data} & \textbf{MNIST-13k} & \textbf{MNIST-100k}  \\
\hline \\
{\sc MHminibatch}   & $129.8\pm 3.4$    & $202.5 \pm 13.6$ \\
{\sc AustereMH(nc)} & $973.8\pm 49.8$   & $1098.3 \pm 44.9$ \\
{\sc AustereMH(c)}  & $1924.3\pm 52.4$  & $2795.6 \pm 364.0$ \\
{\sc MHSubLhd}      & $10783.4\pm 78.9$ & $14977.3 \pm 582.0$ \\
\end{tabular}
\end{center}
\vskip -0.1in
\end{table}

@@

For all methods, we impose a uniform prior on $\theta$ and again use a random
walk proposer, with covariance matrix $0.05I$ for MNIST-13k and $0.01I$ for
MNIST-100k. The default temperature setting is a constant at $K=100$ for
MNIST-13k and MNIST-100k. Performance of all methods implicitly relies on the
step size and temperature. Setting temperature too low or step size too high
will generally result in worse performance for all methods. For MNIST-13k, each
method generated 5000 samples for ten independent trials; due to MNIST-100k's
higher computational requirements, the methods generated 3000 samples for five
independent trials.  For additional parameter settings and an investigation on
tuning step sizes, see Appendix~\ref{app:logistic}.

@@

The first two subplots of Figure~\ref{fig:logistic_performance} display the
prediction accuracy on both datasets for all methods as a function of the
cumulative training points processed.\footnote{The curves do not span the same
length over the x-axis since the methods consume different amounts of data.} To
generate the curves, for each of the sampled vectors $\theta_t$,
$t\in\{1,\ldots,T\}$, we use $\theta_t$ as the logistic regression parameter.
The results indicate that our test is more efficient, obtaining convergence more
than an order of magnitude faster than {\sc AustereMH(nc)} and several orders of
magnitude compared to {\sc AustereMH(c)} and {\sc MHSubLhd}.  We also observe
the advantage of having higher temperature from the third plot in
Figure~\ref{fig:logistic_performance}. During the exploration period, the
accuracy rapidly increases, and then after 200 samples, we switch the
temperature to 1 (equivalent to ignoring temperature). {\rm MHminibatch}
requires the step size to decrease, hence the smaller changes in accuracy.

Figures~\ref{fig:logistic_histograms_mnist}
and~\ref{fig:logistic_histograms_mnist8m} show log-log histograms of minibatch
sizes for the three methods.  The histograms only represent one representative
trial; Table~\ref{tab:logistic} contains the average of the average minibatch
sizes ($\pm$ one standard deviation) across all trials. {\sc MHminibatch}, with
average minibatch sizes of roughly 129.8 and 202.5 for MNIST-13k and MNIST-100k,
respectively, consumes more than 5x fewer data points than the next-best method,
{\sc AustereMH(nc)}.  We reiterate, however, that both {\sc AustereMH(nc)} and
{\sc MHSubLhd} require computing $\log p(x_i|\theta)$ and $\log p(x_i|\theta')$
for all $x_i$ each iteration. Our results here do not count that extra data
    consumption. Only our method and {\sc AustereMH(c)} rely solely on the
    minibatch of data each iteration.

@@

@@

\textbf{Daniel: TODO this is our third experiment. Results are in progress}

@@

\begin{table}[t]
\caption{Parameters for the logistic regression experiments. Unless otherwise
stated, values apply to all methods tested. For the methods
from~\citep{cutting_mh_2014}, we use their notation for clarity.}
\small
\label{tab:logistic_settings}
\vskip 0.15in
\begin{center}
\begin{tabular}{l l l}
\textbf{Value} & \textbf{MNIST-13k} & \textbf{MNIST-100k} \\
\hline
\T Temperature $K$       & 100 & 100 \\
Number of samples $T$ & 5000 & 3000 \\
Number of trials      & 10 & 5 \\
Step size $\sigma$ for random walk proposer with covariance $\sigma I$ & 0.05 & 0.01 \\
{\sc MHminibatch} and {\sc MHSubLhd} minibatch size $m$ & 100 & 100 \\
{\sc AustereMH(c)} $\Delta^*$ bound & 0.1 & 0.2 \\
{\sc AustereMH(c)} minibatch size $m$ from grid search & 450 & 300 \\
{\sc AustereMH(c)} per-test threshold $\epsilon$ from grid search & 0.01 & 0.01\\
{\sc AustereMH(nc)} $\Delta^*$ bound & 0.05 & 0.1 \\
\end{tabular}
\end{center}
\vskip -0.1in
\end{table}

\textbf{Daniel: TODO major work in progress.}

Table~\ref{tab:logistic_settings} indicates the parameter settings for the
logistic regression experiments.

Figure~\ref{fig:logistic_histograms_appendix} indicates minibatch histograms for
one trial of {\sc MHminibatch} on MNIST-100k for proposers $\sigma I$ with
$\sigma \in \{0.005, 0.01, 0.05\}$.

>>>>

% Daniel: NOTE, I'm calling temperature "K" so that "T" stays reserved for
% "number of samples of \theta we draw".
In Sections~\ref{ssec:gaussians} and~\ref{ssec:logistic}, we benchmark {\sc
MHminibatch} against the tests from~\citep{icml2014c1_bardenet14}
and~\citep{cutting_mh_2014}, with the latter using a grid-search over minibatch
sizes $m$ and per-test thresholds $\epsilon$ described in
Appendix~\ref{sssec:grid_search}. Throughout our descriptions, we refer to a
\emph{trial} as the period when an algorithm collects all its desired samples
$\{\theta_1,\ldots,\theta_T\}$, generally with $T=3000$ or $T=5000$.

@@

Figure~\ref{fig:gauss_mix_2} shows that {\sc MHminibatch} dominates in terms of
speed and efficiency. The histograms of the (final) minibatch sizes used each
iteration show that our method consumes significantly less data; the
distribution is short-tailed and the mean is 210, more than an order of
magnitude better compared to the other two methods (averages are 15562 and
16857). Sizes correspond to the running times of the methods, excluding the
likelihood computation of all data points for {\sc MHSubLhd}, which would
drastically increase running time.

@@

% Daniel: moving this here so figures don't overflow to next pages. We can tune
% the positioning if needed.
\begin{figure*}[t]
	\centering
	\includegraphics[width=1.0\linewidth]{logistic_regression_error_regions.png}
	\caption{
    Binary classification accuracy of the MCMC methods on the 1s vs 7s logistic
    regression task for MNIST-13k (left plot) and MNIST-100k (middle plot) as a
    function of cumulative data usage.  The right plot reports performance of
    {\sc MHminibatch} on both datasets when the temperature starts at 100 and
    drops to 1 after a ``burn-in'' period of 200 samples (vertical dashed line)
    of $\theta$.  For all three plots, one standard deviation is indicated by
    the shaded error regions.
    }
	\label{fig:logistic_performance}
\end{figure*}

% Moving here to get positioning better
\begin{figure*}[t]
	\centering
    \includegraphics[width=0.9\linewidth]{logistic_regression_histograms_mnist8m.png}
	\caption{
    Minibatch sizes for a representative trial of logistic regression on
    MNIST-100k (analogous to Figure~\ref{fig:gauss_mix_2}). Both axes are on a
    log scale and have the same ranges across the three histograms.  See
    Section~\ref{ssec:logistic} for details.
    }
	\label{fig:logistic_histograms_mnist8m}
\end{figure*}

@@

For all methods, we impose a uniform prior on $\theta$ and again use a random
walk proposer, with covariance matrix $0.05I$ for MNIST-13k and $0.01I$ for
MNIST-100k. The default temperature setting is a constant at $K=100$ for
MNIST-13k and MNIST-100k. Performance of all methods implicitly relies on the
step size and temperature. Setting temperature too low or step size too high
will result in slow convergence for all methods. For MNIST-13k, each method
generated 5000 samples for ten independent trials; due to MNIST-100k's higher
computational requirements, the methods generated 3000 samples for five
independent trials.  For additional parameter settings and an investigation on
tuning step sizes, see Appendix~\ref{app:logistic}.

@@

The first two subplots of Figure~\ref{fig:logistic_performance} display the
prediction accuracy on both datasets for all methods as a function of the
cumulative training points processed.\footnote{The curves do not span the same
length over the x-axis since the methods consume different amounts of data.} To
generate the curves, for each of the sampled vectors $\theta_t$,
$t\in\{1,\ldots,T\}$, we use $\theta_t$ as the logistic regression parameter.
The results indicate that our test is more efficient, obtaining convergence more
than an order of magnitude faster than {\sc AustereMH(nc)} and several orders of
magnitude compared to {\sc AustereMH(c)} and {\sc MHSubLhd}.  We also observe
the advantage of having higher temperature from the third plot in
Figure~\ref{fig:logistic_performance}. During the exploration period, the
accuracy rapidly increases, and then after 200 samples, we switch the
temperature to 1, but this requires the step size to decrease, hence the smaller
changes in accuracy.

Figure~\ref{fig:logistic_histograms_mnist8m} shows log-log histograms of
minibatch sizes for the three methods on MNIST-100k.
(Figure~\ref{fig:logistic_histograms_mnist} in Appendix~\ref{app:logistic}
contains results for MNIST-13k.) The histograms only represent one
representative trial; Table~\ref{tab:logistic} contains the average of the
average minibatch sizes ($\pm$ one standard deviation) across all trials. {\sc
MHminibatch}, with average minibatch sizes of roughly 129.8 and 202.5 for
MNIST-13k and MNIST-100k, respectively, consumes more than 5x fewer data points
than the next-best method, {\sc AustereMH(nc)}.  We reiterate, however, that
both {\sc AustereMH(nc)} and {\sc MHSubLhd} require computing $\log
p(x_i|\theta)$ and $\log p(x_i|\theta')$ for all $x_i$ each iteration. Our
results here do not count that extra data consumption. Only our method and {\sc
AustereMH(c)} rely solely on the minibatch of data each iteration.

@@

\begin{table}[t]
\caption{Average minibatch sizes ($\pm$ one standard deviation) on logistic
regression on MNIST-13k and MNIST-100k.  The averages are taken over 10
independent trials (5000 samples each) for MNIST-13k and 5 independent trials
(3000 samples each) for MNIST-100k.}
\small % Daniel: yeah, sorry, but otherwise it doesn't fit in a column
\label{tab:logistic}
\vskip 0.15in
\begin{center}
\begin{tabular}{l l l l}
\textbf{Method/Data} & \textbf{MNIST-13k} & \textbf{MNIST-100k}  \\
\hline \\
{\sc MHminibatch}   & $129.8\pm 3.4$    & $202.5 \pm 13.6$ \\
{\sc AustereMH(nc)} & $973.8\pm 49.8$   & $1098.3 \pm 44.9$ \\
{\sc AustereMH(c)}  & $1924.3\pm 52.4$  & $2795.6 \pm 364.0$ \\
{\sc MHSubLhd}      & $10783.4\pm 78.9$ & $14977.3 \pm 582.0$ \\
\end{tabular}
\end{center}
\vskip -0.1in
\end{table}

@@

\textbf{Daniel: TODO this is our third experiment. Results are in progress...}

\textbf{We benchmark with the ADAGrad~\citep{adagrad} implementation of Sparse
Matrix Factorization described in~\citep{Canny:2013:BDA:2487575.2487677}.
(Daniel: is this the correct citation for BIDMach? I assume this is what we will
use)}

@@

\begin{table}[t]
\caption{Parameters for the logistic regression experiments.}
\small
\label{tab:logistic_settings}
\vskip 0.15in
\begin{center}
\begin{tabular}{l l l}
\textbf{Value} & \textbf{MNIST-13k} & \textbf{MNIST-100k} \\
\hline
\T Temperature $K$       & 100 & 100 \\
Number of samples $T$ & 5000 & 3000 \\
Number of trials      & 10 & 5 \\
Step size $\sigma$ for random walk proposer with covariance $\sigma I$ & 0.05 & 0.01 \\
{\sc MHminibatch} and {\sc MHSubLhd} minibatch size $m$ & 100 & 100 \\
{\sc AustereMH(c)} chosen $\Delta^*$ bound & 0.1 & 0.2 \\
{\sc AustereMH(c)} minibatch size $m$ from grid search & 450 & 300 \\
{\sc AustereMH(c)} per-test threshold $\epsilon$ from grid search & 0.01 & 0.01\\
{\sc AustereMH(nc)} chosen $\Delta^*$ bound & 0.05 & 0.1 \\
{\sc MHSubLhd} $\gamma$ & 2.0  & 2.0  \\
{\sc MHSubLhd} $p$      & 2 & 2 \\
{\sc MHSubLhd} $\delta$ & 0.01 & 0.01 \\
\end{tabular}
\end{center}
\vskip -0.1in
\end{table}

Figure~\ref{fig:logistic_histograms_mnist} shows the histograms for the four
methods on one representative trial of MNIST-13k, which shows similar relative
performance of the four methods as shown in
Figure~\ref{fig:logistic_histograms_mnist8m} (which uses MNIST-100k). In
particular, {\sc MHminibatch} exhibits a shorter-tailed distribution and
consumes nearly an order of magnitude fewer data points compared to {\sc
AustereMH(nc)}, the next-best method; see Table~\ref{tab:logistic} for details.

Next, we investigate the impact of the step size $\sigma$ for the random walk
proposers with covariance matrix $\sigma I$. Note that $I$ is $784\times 784$ as
we did not perform any downsampling or data preprocessing other than rescaling
the pixel values to lie in $[0,1]$.

For this, we use the larger dataset MNIST-100k, and test with $\sigma \in
\{0.005, 0.01, 0.05\}$. We keep other parameters consistent with the experiments
in Section~\ref{ssec:logistic}, in particular, keeping the initial minibatch
size $m=100$, which is also the amount the minibatch increments by if we need
more data. Figure~\ref{fig:logistic_histograms_appendix} indicates minibatch
histograms (again, using the log-log scale) for one trial of {\sc MHminibatch}
using each of the step sizes. We observe that by tuning {\sc MHminibatch}, we
are able to adjust the average number of data points in a minibatch across a
wide range of values. Here, the smallest step size results in an average of just
116.1 data points per minibatch, while increasing to $\sigma=0.05$ (the step
size used for MNIST-13k) results in an average of 2215.6. This relative trend is
also present for both {\sc AustereMH} variants and {\sc MHSubLhd}.

Table~\ref{tab:logistic_settings} indicates the relevant parameter settings for
the logistic regression experiments. Unless otherwise stated, values apply to
all methods tested. For values from~\citep{cutting_mh_2014}
or~\cite{icml2014c1_bardenet14}, we use their notation to be consistent.
